\section{The Diffusion Positive Control}
\label{sec:results-diffusion}

The diagnosis of the preceding sections is causal: it attributes the failure of gradient guidance not to the sampler and not to the energy but to the training objective that shapes the energy. An autoregressive model is trained only to predict the next token, and so it is never asked, at any noise level, in which direction the data density increases. A model trained to denoise is asked exactly that, and by the identity that denoising learns the score \citep{vincent2011connection} it acquires the very directional field the autoregressive gradient was found to lack. If the training objective is truly the cause, then a score-trained model on the same tokenizer should provide a usable local direction where the autoregressive gradient did not, and should perform the revision operation that the gradient machinery could not. This section reports that control, run on the score-entropy discrete diffusion (SEDD) model of \citet{lou2024discrete} at two scales, small and medium, which share the GPT-2 tokenizer and drop into the existing masked-recovery harness without retokenization. The results are pilots, on a smaller and differently trained model than the frozen autoregressive backbone, so the comparison is qualitative and directional by design; they are not, and are not read as, a claim that the diffusion model beats the domain-tuned autoregressive baselines in absolute terms.

One clarification governs the comparison throughout, because it is easy to state the contrast too loosely. The relevant difference between the two model families is not that one commits tokens within a pass and the other does not; absorbing diffusion also commits a token at each denoising step and does not thereafter change it. The difference is at the level of the revision \emph{operation}. To revise token $i$, a diffusion model re-masks that position and re-denoises it conditioned on both sides, which is an in-distribution bidirectional query it was trained to answer, whereas an autoregressive model has no conditioning path from the right context to a token at all. It is the availability of that bidirectional conditioning query, not the mere act of committing, that separates the two.

\subsection{The signal: linearization on the unified corpus}
\label{sec:results-diffusion-lin}

The first measurement repeats the linearization diagnostic of Section~\ref{sec:results-linradius} with the diffusion score in place of the autoregressive gradient, on exactly the same sequence set used for the reference baselines, so that the numbers sit on one axis. For each sequence and one masked position, a cheap one-pass surrogate, the model's denoising log-preference for filling the position, is correlated against an expensive per-candidate truth, the measured effect of committing each candidate on the reconstruction of the held-out remainder of the sequence. Table~\ref{tab:diffusion-lin} reports the Spearman correlation, pooled and by embedding-distance stratum. Under the autoregressive objective the surrogate is statistically indistinguishable from zero at every distance, with the magnitude of the correlation below $0.06$ throughout. Under the diffusion objective, on the same tokenizer and the same four hundred thousand candidate pairs, the correlation is clearly positive in every stratum and at both scales, with a pooled value near $0.13$ and a per-sequence mean near $0.18$, and the medium model exceeds the small one in the near stratum, the regime where a first-order surrogate is most expected to apply. On a second corpus of narrative text the same measurement gave a pooled correlation of $0.184$ and a near-stratum correlation of $0.306$; the near-stratum value is corpus-dependent and is not carried over, but the sign and order of magnitude are stable across both corpora. % SOURCE: rev_sedd_lin_small.json, rev_sedd_lin_medium.json (WikiText); rev_sedd_linearization.json (ROCStories); diag_linearization_gpt2sft.json (AR).

\begin{table}[h]
    \centering
    \small
    \begin{tabular}{lccccc}
        \toprule
        Model / objective & ALL & near & mid & random & per-seq mean \\
        \midrule
        Autoregressive gpt2-large (gradient) & $-0.011$ & $+0.027$ & $-0.041$ & $-0.048$ & $-0.011$ \\
        SEDD-small (score) & $+0.130$ & $+0.097$ & $+0.115$ & $+0.126$ & $+0.184$ \\
        SEDD-medium (score) & $+0.133$ & $+0.148$ & $+0.122$ & $+0.110$ & $+0.179$ \\
        \bottomrule
    \end{tabular}
    % SOURCE: spearman_surrogate_vs_true by stratum. AR diag_linearization_gpt2sft.json;
    % SEDD-small rev_sedd_lin_small.json; SEDD-medium rev_sedd_lin_medium.json. n=200,
    % 400,000 candidate pairs, sigma 0.1, same WikiText-2 sequence set as the baselines.
    \caption{Linearization correlation between the cheap one-pass surrogate and the expensive per-candidate truth, on the unified WikiText-2 sequence set ($n=200$, four hundred thousand candidate pairs). The autoregressive gradient is statistically zero at every embedding distance; the diffusion score is clearly positive in every stratum at both scales. This is the signal-level confirmation of the training-objective diagnosis.}
    \label{tab:diffusion-lin}
\end{table}

\subsection{The capability: native recovery of a flipped token}
\label{sec:results-diffusion-recovery}

A usable local direction is a necessary condition for the capability the gradient machinery lacked, but not obviously a sufficient one, so the second measurement asks the diffusion model to perform the operation itself. On the same corrupted sequences the reference baselines used, the flipped token is re-masked and the position is re-denoised conditioned on both sides, a single native draw with no reranking and no sampling of many candidates. Table~\ref{tab:diffusion-recovery} reports exact-match recovery, top-five recovery, and the KL divergence under the autoregressive model that the whole chapter uses as its yardstick. Both scales recover the exact token in roughly a third of cases and place the true token in the top five two-thirds of the time, and the resulting KL, near $3.7$, sits well below the untouched corruption at $9.14$ and well below the flagship Langevin configurations near $6.5$, with the medium model at least matching the small one on every metric. % SOURCE: rev_sedd_recovery_small.json, rev_sedd_recovery_medium.json; reference rows rev_klbase_gpt2sft.json. The comparison to the domain-tuned autoregressive baselines is presented for placement only, since scale and training domain differ; the load-bearing statement is the qualitative one, that a score-trained model natively performs, in a single conditioned pass, the recovery that the gradient-guided sampler could not perform at all.

\begin{table}[h]
    \centering
    \small
    \begin{tabular}{lccc}
        \toprule
        Method & Exact (\%) & Top-5 (\%) & KL under gpt2-large \\
        \midrule
        SEDD-small, native recovery & 32.5 & 61.5 & 3.75 \\
        SEDD-medium, native recovery & 35.0 & 67.0 & 3.73 \\
        \midrule
        \multicolumn{4}{l}{\emph{Reference rows on the same task (autoregressive gpt2-large)}} \\
        Ground-truth fill (floor) & 100 & -- & 0.00 \\
        Top-$k$ rescore (one forward pass) & 33 & -- & 4.43 \\
        Gibbs (gradient-free) & 18.5 & -- & 6.69 \\
        Flagship Langevin & 0.0 & -- & $\approx 6.5$ \\
        Untouched corruption (ceiling) & 0.0 & -- & 9.14 \\
        \bottomrule
    \end{tabular}
    % SOURCE: rev_sedd_recovery_{small,medium}.json by_arm.sedd_recovery;
    % reference rows rev_klbase_gpt2sft.json by_baseline.*.
    \caption{Native diffusion recovery of one flipped token on the reference sequence set, against autoregressive reference rows on the identical task. The diffusion KL sits well below the untouched corruption and the Langevin configurations. The reference rows are for placement only; scale and training domain differ, so the comparison is qualitative.}
    \label{tab:diffusion-recovery}
\end{table}

\subsection{Sufficiency: the repaired signal inside the thesis's own machinery}
\label{sec:results-diffusion-hybrid}

The recovery result changes the model, the sampler, and the energy all at once, so it does not by itself isolate the direction signal as the repaired component. A cleaner test holds the thesis's own machinery fixed and swaps in only the direction. The hybrid sampler runs a position-wise independence Metropolis chain on the same sequences and positions, accepting by the \emph{exact} autoregressive sequence-log-likelihood energy used everywhere in this chapter, and changes exactly one thing: the proposal distribution is the diffusion model's one-pass bidirectional log-preference over the vocabulary rather than the autoregressive left-context conditional. Table~\ref{tab:diffusion-hybrid} reports the outcome against three reference arms run on the identical task: the autoregressive left-conditional independence sampler, and the discrete Langevin policy and random arms.

Swapping only the direction signal, inside the same Metropolis-corrected chain, on the same exact energy and the same task, turns a sampler that never recovers into one that works. The gradient arm recovers the token in $0.0\%$ of cases, the autoregressive left-conditional proposal in $23.5\%$, and the diffusion-proposal hybrid in $38.5\%$ at the small scale and $39.0\%$ at the medium scale, with acceptance rising from $34.4\%$ for the left-conditional to above $40\%$ for the hybrid because the bidirectional proposal, seeing both sides of the mask, more often lands on tokens the exact energy will accept. % SOURCE: rev_sedd_hybrid.json by_arm.{hybrid_small,hybrid_medium,left_conditional,dls_policy,dls_random}. That the hybrid also beats the left-conditional, and not only the gradient arm, is the informative part: it measures what the bidirectional conditioning adds beyond an exact reweighting of the left context alone. Read alongside the last-token result, the two halves of the argument close. The zero-gradient theorem of Section~\ref{sec:results-lasttoken} is the necessity half, that a usable direction cannot be extracted from the autoregressive input-embedding gradient; this hybrid is the sufficiency half, that a usable direction, once supplied by a score-trained model, works inside the exact machinery the thesis built.

\begin{table}[h]
    \centering
    \small
    \begin{tabular}{lcccc}
        \toprule
        Sampler (same exact energy) & Proposal & Exact (\%) & Top-5 (\%) & Accept (\%) \\
        \midrule
        Hybrid, SEDD-medium & bidirectional score & 39.0 & 67.0 & 43.3 \\
        Hybrid, SEDD-small & bidirectional score & 38.5 & 61.5 & 40.0 \\
        Left-conditional independence & AR left context & 23.5 & 34.5 & 34.4 \\
        Discrete Langevin, policy & AR gradient & 0.0 & 34.5 & -- \\
        Discrete Langevin, random & random direction & 0.0 & 34.5 & -- \\
        \bottomrule
    \end{tabular}
    % SOURCE: rev_sedd_hybrid.json by_arm.* (exact_match_pct, top5_pct, mean_accept_pct).
    \caption{The hybrid sampler, which changes only the proposal direction inside the thesis's exact-energy Metropolis chain on the reference sequence set. Replacing the autoregressive left-context proposal ($23.5\%$ exact) or the gradient arm ($0.0\%$) with the diffusion model's bidirectional one-pass preference lifts exact recovery to $38.5\%$ to $39.0\%$. This is the sufficiency half of the diagnosis.}
    \label{tab:diffusion-hybrid}
\end{table}

\subsection{Trained guidance: steering with a noisy-state classifier}
\label{sec:results-diffusion-guided}

The final pilot asks whether the constructive direction extends to control, the capability the constraint gradient of Section~\ref{sec:results-constrained} did not reliably provide. A small classifier is trained to read a noisy, partially masked state, the state the diffusion model occupies mid-denoising, corrupting each labelled example with the model's own forward absorbing process at a noise level drawn uniformly over the schedule. Its held-out accuracy at low noise is $81\%$, well above chance, and it degrades smoothly as noise rises.
% SOURCE: noisy_classifier_train.json final_low_noise_acc=80.96.
During generation this classifier reweights the diffusion model's top thirty-two candidates at each commitment by the classifier's probability of the target label, raised to a guidance strength $\gamma$, and it only guides. Steering is scored, exactly as in Section~\ref{sec:results-constrained}, by the frozen-model sentiment judge, so the classifier that guides and the judge that grades are strictly separate. The fairness of the comparison to the autoregressive constraint arms is that those too used a trained classifier; the only difference is that this one is trained on the noisy states it is evaluated on, which is the thesis's own principle applied in the constructive direction.

Guidance is examined in two settings that differ only in whether the guiding and the judging classifier are calibrated to the text being scored, because that difference turns out to govern how far the steering transfers. The first is an off-domain continuation run, in which prompts come from a MuCoLa-style continuation set neither classifier was trained on; it is reported immediately below. The second, reported after it as a control on the first, uses held-out in-domain prompts on which both classifiers are calibrated and adds a fluency trust region that caps the per-commitment departure from the diffusion model's own preference, so that the steering can be read without the confound of collapsing fluency.

In the off-domain setting, Table~\ref{tab:diffusion-guided} reports the result, and it is honest to state it as it came out rather than as the prediction hoped. The guidance mechanism demonstrably works: it moves the guiding classifier's own verdict toward the target at every strength and for both labels, by seven to eighteen points. Measured by the independent judge, it steers significantly and monotonically in the negative direction, from a six-point gain at $\gamma=1$ to a $16.7$-point gain at $\gamma=4$ whose confidence interval excludes zero, but it does not move the judge in the positive direction. The asymmetry has a clear cause that the numbers isolate. The guiding classifier and the judge agree on only fifty-six to sixty-four percent of the generated texts, because those texts are off the fluent manifold on which both classifiers were trained, and stronger guidance pushes them further off it, with the per-token negative log-likelihood under the autoregressive model rising from $7.1$ to $11.0$ as $\gamma$ grows. % SOURCE: rev_sedd_guided_g{1,2,4}.json by_label.*. Guiding hard toward one classifier's notion of a label therefore need not raise the other's, precisely the off-manifold breakdown that Section~\ref{sec:results-constrained} observed for the constraint-only arm. The pilot thus shows both sides of the constructive claim: trained noisy-state guidance is a working steering mechanism, and its reach on an independent judge is bounded by the same off-manifold fragility of classifiers that runs throughout the thesis, at a fluency cost that grows with the guidance strength.

\begin{table}[h]
    \centering
    \small
    \begin{tabular}{lccccc}
        \toprule
        & \multicolumn{2}{c}{Judge target-hit (\%)} & Judge gain & Self gain & NLL/token \\
        \cmidrule(lr){2-3}
        $\gamma$, target & unguided & guided & (pts, [CI]) & (pts) & (un$\to$guided) \\
        \midrule
        $1$, negative & 49 & 55 & $+6.0\ [-1.3,13.3]$ & $+9.7$ & $7.1\to8.5$ \\
        $1$, positive & 44 & 39 & $-5.3\ [-12.3,1.7]$ & $+11.0$ & $7.2\to8.5$ \\
        $2$, negative & 49 & 60 & $+10.7\ [3.3,18.0]$ & $+7.7$ & $7.1\to9.7$ \\
        $2$, positive & 44 & 44 & $+0.0\ [-7.0,7.0]$ & $+12.0$ & $7.2\to9.6$ \\
        $4$, negative & 49 & 66 & $+16.7\ [9.7,24.0]$ & $+7.3$ & $7.1\to11.0$ \\
        $4$, positive & 44 & 39 & $-5.3\ [-13.0,2.7]$ & $+18.3$ & $7.2\to11.0$ \\
        \bottomrule
    \end{tabular}
    % SOURCE: rev_sedd_guided_g{1,2,4}.json by_label.{0,1}: unguided_hit_pct, guided_hit_pct,
    % gain_pts + gain_ci95, self_gain_pts, unguided_span_nll -> guided_span_nll.
    \caption{Classifier-guided diffusion generation on the \emph{off-domain} continuation task, scored by the independent frozen-model judge (never by the guiding classifier). The guidance moves the guiding classifier's own verdict for both labels (self gain), and moves the independent judge significantly in the negative direction (confidence interval excludes zero at $\gamma\ge2$) but not the positive direction. The asymmetry tracks the off-manifold disagreement between the two classifiers and a fluency cost, in per-token negative log-likelihood, that grows with the guidance strength. The on-domain trust-region control of Table~\ref{tab:gprime} removes the fluency cost and flips the working direction.}
    \label{tab:diffusion-guided}
\end{table}

\subsection{On-domain guidance under a fluency trust region}
\label{sec:results-diffusion-gprime}

The off-domain result leaves two explanations entangled. The guidance might steer only in one fixed direction, or the two independently trained classifiers might simply disagree on the off-domain text they were never calibrated for, with stronger guidance driving the text further off the fluent manifold where the disagreement is worst. The continuation run cannot separate these, because it varies domain and fluency together. The control that separates them holds fluency fixed and moves the instruments on-domain. Prompts are the first three hundred held-out SST-2 validation sentences of at least twelve tokens; neither the guiding classifier nor the judge trained on this split, both having only read it under \texttt{no\_grad} for an accuracy readout, so the split is calibration-clean by construction. A trust region caps guidance to candidates within $\delta=5$ nats of the diffusion model's top choice at each commitment, so a committed token is always within $\delta$ nats of the model's own preference and guidance can only reshuffle mass among near-equally-fluent options. Unguided and guided generations share a seed per prompt, so each pair differs only by the intervention.

The first thing the control establishes is that the instrument diagnosis of the off-domain run was correct. The agreement between the guiding classifier and the judge climbs a clean ladder as the text they score moves on-domain: from the off-domain band of $56$ to $64\%$, to $71.7\%$ (95\% CI $[66.7, 76.7]$) on the unguided on-domain generations, to $79.7\%$ ($[75.0, 84.0]$) on the real held-out sentences themselves, where the judge and the guide reach $88\%$ and $79.7\%$ accuracy against the gold labels.
% SOURCE: results_revision/rev_gprime.json diagnosis_unguided_gen_agreement_pct 71.667 [66.667,76.667]; realtext_agreement_pct 79.667 [75,84]; realtext_judge_acc_pct 88.0; realtext_clf_acc_pct 79.667; phase4_offdomain_agreement_band_pct [56,64].
The two classifiers are therefore not miscalibrated in some fixed way; they simply agree more the closer the scored text sits to the distribution they were trained on, which is exactly what the off-domain diagnosis predicted.

With fluency held fixed by the trust region, the steering itself is reported in Table~\ref{tab:gprime}. Three facts stand out. The trust region does its job: guided span negative log-likelihood stays at or below the unguided value in three of the four cells and rises by at most $0.26$ nats in the fourth, eliminating the $7.1\to11.0$ climb of the off-domain run entirely.
% SOURCE: rev_gprime.json by_cell.*.nll_cost (-0.74, -1.71, +0.26, -1.00); unguided_span_nll 7.014.
The guide steers its own verdict strongly in both directions at every cell, by $+9.3$ to $+25.7$ points with confidence intervals excluding zero, so the mechanism moves both ways.
% SOURCE: rev_gprime.json by_cell.*.self_gain_pts and self_gain_ci95.
But transfer to the independent judge is one-directional: the positive target gains $+6.7$ and $+7.3$ points at $\gamma=2$ and $\gamma=4$, both intervals excluding zero, while the negative target is null at both strengths, its interval straddling zero.
% SOURCE: rev_gprime.json by_cell.*.gain_pts and gain_ci95.

\begin{table}[h]
    \centering
    \small
    \begin{tabular}{llccccc}
        \toprule
        $\gamma$ & target & unguided & guided & judge gain [CI] & self gain & NLL (un$\to$guided) \\
        \midrule
        $2$ & negative & $53.0$ & $50.7$ & $-2.3\ [-7.0, 2.3]$   & $+18.0$ & $7.01\to6.27$ \\
        $2$ & positive & $47.0$ & $53.7$ & $+6.7\ [1.7, 11.7]$   & $+9.3$  & $7.01\to5.30$ \\
        $4$ & negative & $53.0$ & $52.3$ & $-0.7\ [-5.3, 4.0]$   & $+25.7$ & $7.01\to7.28$ \\
        $4$ & positive & $47.0$ & $54.3$ & $+7.3\ [2.7, 12.0]$   & $+17.7$ & $7.01\to6.01$ \\
        \bottomrule
    \end{tabular}
    % SOURCE: results_revision/rev_gprime.json by_cell.{gamma2_label0,gamma2_label1,
    % gamma4_label0,gamma4_label1}: unguided_hit_pct, guided_hit_pct, gain_pts+gain_ci95,
    % self_gain_pts, unguided_span_nll -> guided_span_nll. n=300 pairs per cell.
    \caption{On-domain, trust-region guided diffusion generation on held-out SST-2 prompts ($n=300$ pairs per cell), scored by the independent judge. Guided target-hit percentage against the paired unguided baseline, with a paired bootstrap $95\%$ confidence interval; the self gain is the guiding classifier's own verdict; the last column is the span negative log-likelihood before and after guidance. The trust region keeps fluency at or below the unguided level, and the judge gain is positive with an interval excluding zero for the positive target at both strengths but null for the negative target. Compare the off-domain run of Table~\ref{tab:diffusion-guided}, where the working direction was the opposite.}
    \label{tab:gprime}
\end{table}

The working direction has therefore flipped relative to the off-domain run: there the judge moved on the negative target and not the positive, here on the positive and not the negative, with fluency no longer a confound. This flip is the decisive evidence about what the residual limit is and is not. It is not a fluency artifact, because the trust region removed the fluency cost and the asymmetry remained. And it is not a fixed unreachable sentiment, because the reachable direction moved when the setting changed. What remains is an instrument-alignment effect: the two independently trained classifiers still disagree on about $28\%$ of the on-domain generations, and steering toward one classifier's notion of a label registers on the other only where the two happen to align on the generated text. The claim the data licenses is therefore setting-contingent and deliberately narrow: with a single noisy guide at this scale, one-directional transfer to an independent judge is reachable under bounded fluency, but symmetric transfer at both labels is not, and which direction transfers tracks instrument alignment rather than a general ``positive steers, negative does not'' rule.

It is tempting to explain the flip by supposing the residual disagreement falls asymmetrically by class, so that the guide's positive verdict transfers to the judge more reliably than its negative verdict. Table~\ref{tab:confusion} tests that supposition directly on the neutral calibration surface, the unguided on-domain generations, and does not support it. The per-class accuracies rule out a silent label-map flip: the guide is $79.4\%$ accurate on negative and $79.9\%$ on positive held-out sentences, the judge $85.8\%$ and $89.9\%$, both balanced.
% SOURCE: results_revision/rev_confusion.json label_alignment_sanity.
On the unguided generations the judge confirms the guide's positive verdict $67.9\%$ of the time and its negative verdict $75.7\%$, a paired difference of $-7.7$ points whose $95\%$ confidence interval $[-18.0, +2.6]$ includes zero and in fact leans, if at all, the other way.
% SOURCE: results_revision/rev_confusion.json confusion_unguided_ondomain: P_judge_pos_given_guide_pos 67.95 [60.4,75.3], P_judge_neg_given_guide_neg 75.69 [68.5,82.4], conditional_diff -7.75 [-18.0,2.6].
The two conditional agreement rates are statistically indistinguishable, so the confusion table confirms the overall alignment gap that the ladder measures but does not identify a by-class asymmetry that would by itself explain the transfer direction. The honest reading is that the transfer asymmetry is real and observed but its class-level cause is not resolved by this data, and no further mechanism is asserted beyond the alignment gap. Representative guided pairs from both settings, drawn by a fixed seed and shown unfiltered, are collected in Appendix~\ref{app:gprime-examples}, including the off-domain high-$\gamma$ pairs whose fluency the trust region repairs.

\input{chapters/tab_confusion}

Taken together, these four pilots move the diffusion control from a proposed experiment to a completed one. The signal the autoregressive gradient lacked is present in the diffusion score; the recovery the gradient machinery could not perform is performed natively; the repaired direction works inside the thesis's own exact-energy chain; and trained guidance steers, with a reach limited by the same off-manifold fragility the thesis documents elsewhere. The training objective is thereby isolated as the causal variable, at both the signal level and the capability level.
