\documentclass[12pt,leqno,a4paper]{report}

% ---- packages (kept close to the IMS template the department provides) ----
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{longtable} % multi-page guided-generation example tables in the appendix
\usepackage{multirow} % two-by-two confusion table in the diffusion-control section
\usepackage{amsmath,amssymb}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage[paper=a4paper,left=3cm,right=3cm,top=2.5cm,bottom=2.5cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[german,english]{babel} % english is the main language (last option); german loaded for the Erklaerung, matching the IMS template which loads both
\usepackage{xcolor}
\usepackage[draft=false,protrusion=false,expansion=false]{microtype}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{xurl}      % break long URLs anywhere so bibliography entries do not overflow the margin (matches Ioanna's preamble)
\urlstyle{same}
\usepackage{fancyhdr}
\usepackage{setspace}
\usepackage{titlesec} % compact chapter headings ("5  Methods") in place of the large "Chapter 5" banner, matching the department template's section-style headings
% Render chapter headings on a single line: the number, then the title, no "Chapter" word and no oversized vertical block.
\titleformat{\chapter}[hang]{\normalfont\huge\bfseries}{\thechapter}{0.6em}{}
\titlespacing*{\chapter}{0pt}{0pt}{18pt}
\renewcommand{\chaptermark}[1]{\markboth{\thechapter.\ #1}{}} % compact running header ("5. Results"), no "CHAPTER" banner

\renewcommand{\baselinestretch}{1.3}
\parskip=\medskipamount
\frenchspacing
\bibpunct[; ]{(}{)}{;}{a}{,}{;} % citation punctuation pulled verbatim from the official IMS template (Schweitzer/Reiter)

% ---- notation shorthands used throughout ----
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\vs}{\bm{s}}
\newcommand{\ve}{\bm{e}}
\newcommand{\vg}{\bm{g}}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\energy}{E}
\newcommand{\grad}{\nabla}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator{\KL}{KL}

\captionsetup{font=small,labelfont=bf,width=0.95\textwidth}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\leftmark}
\fancyhead[R]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

\begin{document}

% ============================================================
% TITLE PAGE
% ============================================================
\thispagestyle{empty}
\begin{center}
    {\large Institut f\"ur Maschinelle Sprachverarbeitung}\\
    {\large Universit\"at Stuttgart}\\
    {\large Pfaffenwaldring 5B}\\
    {\large D-70569 Stuttgart}\\[3cm]

    {\large Master Thesis}\\[1.2cm]

    {\LARGE\bfseries Energy-Guided Sampling for Controllable Text Generation: From Langevin Dynamics to Amortized Inference}\\[2cm]

    {\large Sarthak Singh}\\[2cm]

    \begin{tabular}{ll}
        Studiengang: & M.Sc. Computational Linguistics\\[0.6cm]
        Pr\"ufer*innen: & Prof. Dr. Ngoc Thang Vu\\
                        & \\[0.6cm] % AUTHOR DECISION: enter the second examiner (two examiners are required by the IMS guidelines)
        Betreuer*in: & Prof. Dr. Ngoc Thang Vu\\[1.2cm]
        Beginn der Arbeit: & \\ % AUTHOR DECISION: enter start date (dd.mm.yyyy)
        Ende der Arbeit: & \\ % AUTHOR DECISION: enter end date (dd.mm.yyyy)
    \end{tabular}
\end{center}
\clearpage

% ============================================================
% STATEMENT OF AUTHORSHIP (newer wording, with AI-tools clause)
% ============================================================
\thispagestyle{empty}
\selectlanguage{german}
\noindent\textbf{Erkl\"arung (Statement of Authorship)}\\[0.4cm]

\noindent Hiermit erkl\"are ich, dass ich die vorliegende Arbeit selbstst\"andig verfasst und dabei keine anderen als die angegebenen Quellen und Hilfsmittel benutzt habe. S\"amtliche w\"ortlichen oder sinngem\"a\ss{}en \"Ubernahmen und Zitate aus anderen Werken sind kenntlich gemacht und nachgewiesen. Ich versichere, dass ich keine Hilfsmittel verwendet habe, deren Nutzung die Pr\"uferin oder der Pr\"ufer explizit ausgeschlossen hat. Im Anhang \glqq Nutzung von KI-Tools\grqq{} habe ich dokumentiert, welche KI-Tools zur Erstellung wesentlicher Teile dieser Arbeit in welcher Art und Weise verwendet wurden. Mit Abgabe der Arbeit \"ubernehme ich die volle Verantwortung f\"ur alle Inhalte. Die Grunds\"atze zur KI-Nutzung f\"ur Studierende an der Universit\"at Stuttgart habe ich zur Kenntnis genommen und befolgt. Weder diese Arbeit noch wesentliche Teile daraus waren bisher Gegenstand eines anderen Pr\"ufungsverfahrens. Das elektronische Exemplar stimmt mit allen eingereichten Exemplaren \"uberein.\footnote{\foreignlanguage{english}{Non-binding translation for convenience: I hereby declare that the work presented in this thesis is entirely my own. I did not use any other sources and aids than the listed ones. I have marked all direct or indirect transfers from other sources contained therein as referenced quotations. I assure that I have not used any aids that were explicitly ruled out by the examiner. In the Appendix \glqq Use of AI Tools\grqq{} I have documented which AI tools were used and how they contributed in the creation of this work. By submitting this work, I assume full responsibility for all content. I have taken notice of and I comply with the Principles for the Use of Artificial Intelligence by Students at the University of Stuttgart. Neither this work nor significant parts of it were part of another examination procedure. The electronic copy is consistent with all submitted hard copies.}}\\[2cm]
\selectlanguage{english}

\noindent Stuttgart, \today\\[1.5cm] % AUTHOR DECISION: replace \today with the actual submission date; the printed copy is to be signed by hand above the name line
\noindent (Sarthak Singh)
\clearpage

% ============================================================
% ABSTRACT
% ============================================================
\input{chapters/abstract}

% ============================================================
% FRONT MATTER
% ============================================================
\pagenumbering{roman}
\setcounter{page}{1}
\tableofcontents
\clearpage
\listoffigures
\clearpage
\listoftables
\clearpage

% ============================================================
% BODY
% ============================================================
\pagenumbering{arabic}
\setcounter{page}{1}

\input{chapters/01_introduction}
\input{chapters/02_background}
\input{chapters/03_related_work}
\input{chapters/04_methodology}
\input{chapters/05_results}
\input{chapters/06_discussion}
\input{chapters/07_conclusion}

% ============================================================
% BIBLIOGRAPHY
% ============================================================
\bibliographystyle{plainnat}
\bibliography{references}

% ============================================================
% APPENDIX
% ============================================================
\appendix
\input{chapters/08_appendix}

\end{document}
